---
title: "Disentangling Ability and Non-response in Adaptive Learning Systems: An IRTree Analysis of Problem-Skipping in the Prowise Learn Series Game"
author: "Annie Johansson"
date: "2024-08-15"
output: html_document
---


```{r source-config, include=FALSE}
# Source config via absolute path so the working directory need not be set
# beforehand. Mirrors the detection logic in 00_config.R.
# Priority: Rscript CLI -> RStudio -> source() call -> working directory.
.this_dir <- (function() {
  args <- commandArgs(trailingOnly = FALSE)
  file_arg <- grep("--file=", args, value = TRUE)
  if (length(file_arg) > 0) {
    return(dirname(normalizePath(sub("--file=", "", file_arg))))
  }
  if (
    requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()
  ) {
    ctx <- rstudioapi::getSourceEditorContext()
    if (!is.null(ctx) && nchar(ctx$path) > 0) {
      return(dirname(normalizePath(ctx$path)))
    }
  }
  src <- tryCatch(
    dirname(normalizePath(sys.frames()[[1]]$ofile)),
    error = function(e) NULL
  )
  if (!is.null(src)) {
    return(src)
  }
  getwd()
})()
source(file.path(.this_dir, "00_config.R"))
```

# Load data

```{r prep-data}
# Source the prep script to load data and apply cleaning
source(file.path(.this_dir, "02_irtree_prep.R"))

# Build the full-sample wide table for descriptive analyses.
# The raw data object was already loaded inside 02_irtree_prep.R, so it is
# available via get(). Grade filter and derived columns are applied here to
# match the cleaned data in the modelling pipeline.
logs <- as.data.table(get(data_object))
logs[, question_mark := as.integer(answer == "¿")]
logs[,
  response_type := ifelse(
    question_mark == 1,
    "skip",
    ifelse(correct_answered == 0, "error", "correct")
  )
]
```

```{r check-exclusion}
# Flag which users in the full sample pass the inclusion criterion (min_skips).
logs[,
  in_model := factor(
    user_id %in% dat_irt[, unique(user_id)],
    labels = c("Excluded", "Included")
  )
]
p_exclusion <- ggplot(logs[, .(count_skip = sum(question_mark)), .(in_model, user_id)]) +
  geom_histogram(
    aes(x = count_skip, fill = in_model, color = in_model),
    alpha = 0.7,
    bins = 50
  ) +
  scale_fill_manual(values = c('grey70', 'grey20')) +
  scale_color_manual(values = c('grey70', 'grey20')) +
  labs(
    title = "Histogram of Total Number of Skips per Learner",
    x = "Number of Skips",
    y = "Frequency"
  ) +
  theme_bw()
p_exclusion
save_plot(p_exclusion, name = "skip_distribution_exclusion")
```


# Descriptives

This section describes the raw patterns in skipping before any modelling. The goal is to characterise how problem-skipping varies across difficulty levels, grades, response time, and session progression — and to identify whether there are systematic patterns that motivate the use of a multidimensional measurement model.

## Skip rate by difficulty level

Items in the Series game are assigned to one of three difficulty levels (Easy, Medium, Hard), targeting correct response rates of approximately 90%, 75%, and 60% respectively. We expect the skip rate to increase with difficulty: harder items should elicit more skipping, as students are less likely to know the answer and more likely to avoid the risk of a negative score.

```{r plot-difficulty}
logs[,
  diff_factor := factor(
    difficulty,
    levels = (0:2),
    labels = c("Easy", "Medium", "Hard")
  )
]
logs_diff <- logs[,
  .(
    n_skip = sum(question_mark),
    mean_skip = mean(question_mark),
    sd_skip = sd(question_mark),
    ci_l = t.test(question_mark)$conf.int[1],
    ci_u = t.test(question_mark)$conf.int[2]
  ),
  diff_factor
][order(diff_factor)]

logs_diff

p_difficulty <- ggplot(logs_diff, aes(x = diff_factor, y = mean_skip)) +
  geom_bar(stat = "identity", fill = "gray") +
  geom_errorbar(
    aes(ymin = ci_l, ymax = ci_u),
    color = "gray30",
    width = 0.1
  ) +
  labs(x = "Difficulty Level", y = "Proportion Skip", title = "Proportion of Skipping Across Difficulty Level") +
  theme_bw() +
  theme(
    text = element_text(size = 14)
  )
p_difficulty
save_plot(p_difficulty, name = "skip_by_difficulty")
```

I also do a chi-square test, with Cramer's V, to quickly check the significance of these differences. 

```{r test-difficulty}
ct_diff <- chisq.test(table(logs$diff_factor, logs$question_mark))
cat(sprintf(
  "Chi-square test (skip ~ difficulty): χ²(%d) = %.2f, p %s\n",
  ct_diff$parameter,
  ct_diff$statistic,
  ifelse(ct_diff$p.value < .001, "< .001", sprintf("= %.3f", ct_diff$p.value))
))
cat(sprintf(
  "Cramér's V = %.3f\n",
  sqrt(ct_diff$statistic / (sum(ct_diff$observed) * (min(dim(ct_diff$observed)) - 1)))
))
```

## Skip rate by grade

Across grades 3–8, we expect a small overall increase in QM usage, as older students may be more aware of the scoring implications of a skip response. A chi-square test checks for an overall association between grade and skip rate; a logistic regression with grade as a continuous predictor tests for a linear trend and gives the odds ratio per one-grade increase.

```{r plot-grade}
logs_grade <- logs[,
  .(
    n_skip = sum(question_mark),
    mean_skip = mean(question_mark),
    sd_skip = sd(question_mark),
    ci_l = t.test(question_mark)$conf.int[1],
    ci_u = t.test(question_mark)$conf.int[2]
  ),
  .(grade)
][order(grade)]
logs_grade$grade <- factor(logs_grade$grade, levels = c(3:8))
logs_grade
p_grade <- ggplot(logs_grade, aes(x = grade, y = mean_skip)) +
  geom_path(group = 1, color = "gray20", linewidth = 1) +
  geom_errorbar(aes(ymin = ci_l, ymax = ci_u), color = "gray20", width = 0.1) +
  labs(x = "Grade", y = "Proportion Skip") +
  theme_bw() +
  theme(
    text = element_text(size = 14)
  )
p_grade
save_plot(p_grade, name = "skip_by_grade")
```

I perform a chi-square test to inspect the overall association between grade and skip rate. I also perform a logistic regression to check for a linear trend across grades. 

```{r test-grade}
# Chi-square test for an overall association between grade and skip rate.
ct_grade <- chisq.test(table(logs$grade, logs$question_mark))
cat(sprintf(
  "Chi-square test (skip ~ grade): χ²(%d) = %.2f, p %s\n",
  ct_grade$parameter,
  ct_grade$statistic,
  ifelse(ct_grade$p.value < .001, "< .001", sprintf("= %.3f", ct_grade$p.value))
))
cat(sprintf(
  "Cramér's V = %.3f\n",
  sqrt(ct_grade$statistic / (sum(ct_grade$observed) * (min(dim(ct_grade$observed)) - 1)))
))

# Logistic regression to test for a linear trend across grades.
# Grade is treated as continuous; the coefficient gives the change in log-odds
# of skipping per one-grade increase.
m_grade     <- glm(question_mark ~ grade, data = logs, family = binomial)
m_grade_sum <- summary(m_grade)
b  <- coef(m_grade_sum)["grade", "Estimate"]
se <- coef(m_grade_sum)["grade", "Std. Error"]
z  <- coef(m_grade_sum)["grade", "z value"]
p  <- coef(m_grade_sum)["grade", "Pr(>|z|)"]
cat(sprintf(
  "\nLinear trend (logistic): b = %.3f, SE = %.3f, z = %.2f, p %s\n",
  b, se, z,
  ifelse(p < .001, "< .001", sprintf("= %.3f", p))
))
cat(sprintf("Odds ratio per grade increase: %.3f\n", exp(b)))
```

## Skip rate by grade and difficulty

When grade and difficulty are considered simultaneously, the relationship changes. A likelihood ratio test compares a logistic regression with main effects only to one including the grade × difficulty interaction term, formally testing whether the grade slope differs across difficulty levels. The interaction coefficients give the change in the grade slope for Medium and Hard items relative to Easy.

```{r plot-difficulty*grade}
logs_grade_diff <- logs[,
  .(
    n_skip = sum(question_mark),
    mean_skip = mean(question_mark),
    ci_l = t.test(question_mark)$conf.int[1],
    ci_u = t.test(question_mark)$conf.int[2]
  ),
  .(grade, difficulty = diff_factor)
][order(grade, difficulty)]

logs_grade_diff$grade <- factor(logs_grade_diff$grade, levels = c(3:8))

p_grade_diff <- ggplot(
  logs_grade_diff,
  aes(
    x = grade,
    y = mean_skip,
    fill = difficulty,
    group = difficulty,
    color = difficulty
  )
) +
  geom_ribbon(
    aes(ymin = ci_l, ymax = ci_u),
    color = NA,
    alpha = 0.3,
    width = 0.1
  ) +
  geom_path(linewidth = 1) +
  scale_discrete_manual(
    aesthetics = c("fill", "color"),
    values = colors[c(3, 4, 2)],
    name = "Difficulty Level"
  ) +
  labs(x = "Grade", y = "Proportion Skip") +
  theme_bw() +
  theme(
    text = element_text(size = 14)
  )
p_grade_diff
save_plot(p_grade_diff, name = "skip_by_grade_difficulty")
```

I test the interaction effect with logistic regression models. 

```{r test-difficulty*grade}
# Logistic regression models with and without the grade × difficulty interaction.
# The likelihood ratio test (LRT) compares model fit to determine whether the
# grade slope differs significantly across difficulty levels.
m_main <- glm(question_mark ~ grade + diff_factor,        data = logs, family = binomial)
m_int  <- glm(question_mark ~ grade * diff_factor,        data = logs, family = binomial)

lrt <- anova(m_main, m_int, test = "Chisq")
cat(sprintf(
  "Likelihood ratio test (grade × difficulty interaction): χ²(%d) = %.2f, p %s\n",
  lrt$Df[2],
  lrt$Deviance[2],
  ifelse(lrt[2, "Pr(>Chi)"] < .001, "< .001", sprintf("= %.3f", lrt[2, "Pr(>Chi)"]))
))

# Interaction coefficients: the grade slope for Medium and Hard relative to Easy
cat("\nInteraction coefficients (log-odds, reference = Easy):\n")
int_coefs <- coef(summary(m_int))
int_rows  <- grep("grade:diff_factor", rownames(int_coefs))
print(round(int_coefs[int_rows, ], 3))

# Exponentiated coefficients for interpretability
cat("\nOdds ratios for interaction terms:\n")
print(round(exp(int_coefs[int_rows, "Estimate"]), 3))
```

## Response type over response time

The response deadline in the Series game is 20 seconds. The timing of a skip response relative to this deadline is informative: a skip early in the response window is consistent with a student quickly deciding they do not know the answer, while a skip close to the deadline may reflect a last resort to avoid a time penalty. We plot the probability of each response type (correct, error, question mark) as a function of response time, overlaid on the distribution of response times to indicate where most observations fall.

```{r plot-rt}
logs[, t := round(response_in_milliseconds / 1000, 1)]

library(DescTools)
logs_t <- logs[t < 20, Freq(response_type), t][order(t)]
time_plot <- ggplot(
  logs_t,
  aes(x = t, y = perc, group = level, color = level)
) +
  geom_line(linewidth = 1) +
  geom_point(aes(size = freq), alpha = 0.6) +
  scale_discrete_manual(
    aesthetics = c("fill", "color"),
    values = colors[c(1, 2, 4)],
    name = "Response",
    labels = c("Correct", "Error", "Question Mark")
  ) +
  geom_vline(xintercept = 20, linetype = "dotted") +
  labs(x = "Response Time", y = "Probability of Response", size = "Frequency") +
  theme_bw() +
  theme(
    text = element_text(size = 14)
  )
time_plot
save_plot(time_plot, name = "rt_response_type")

# Add a histogram of rt to the background of the line plot
time_plot_dist <- ggplot() +
  geom_histogram(
    data = logs[t <= 21],
    aes(x = t, y = after_stat(density)),
    bins = 50,
    color = "gray",
    fill = "gray",
    alpha = 0.4
  ) +
  geom_line(
    data = logs_t,
    aes(x = t, y = perc, group = level, color = level),
    linewidth = 1,
    alpha = 0.6
  ) +
  geom_point(
    data = logs_t,
    aes(x = t, y = perc, group = level, color = level, size = freq),
    alpha = 0.4
  ) +
  scale_discrete_manual(
    aesthetics = c("fill", "color"),
    values = colors[c(1, 2, 4)],
    name = "Response",
    labels = c("Correct", "Error", "Question Mark")
  ) +
  geom_vline(xintercept = 20, linetype = "dotted") +
  labs(x = "Response Time", y = "Probability of Response", size = "Frequency") +
  theme_bw() +
  theme(
    text = element_text(size = 14)
  )

time_plot_dist
save_plot(time_plot_dist, name = "rt_response_type_dist")
```

## Response type over session progression

We expect average skipping and error rates to stabilize over time, due to the adaptivity of the system. 

```{r plot-session}
logs_session <- logs[
  new_user_domain_modified_count > 0 & new_user_domain_modified_count <= 250,
  .(
    mean_skip = mean(question_mark),
    ci_l = t.test(question_mark)$conf.int[1],
    ci_u = t.test(question_mark)$conf.int[2],
    .N
  ),
  new_user_domain_modified_count
][order(new_user_domain_modified_count)]

p_session_skip <- ggplot(logs_session, aes(x = new_user_domain_modified_count, y = mean_skip)) +
  geom_ribbon(aes(ymin = ci_l, ymax = ci_u), alpha = 0.6, fill = colors[4]) +
  geom_path(group = 1, color = colors[4]) +
  geom_point(aes(size = N), color = colors[4], alpha = 0.6) +
  labs(x = "Session Count", y = "Proportion Skip", size = "N") +
  theme_bw() +
  theme(
    text = element_text(size = 14)
  )
p_session_skip
save_plot(p_session_skip, name = "skip_over_sessions")

logs_session <- logs[
  new_user_domain_modified_count > 0 & new_user_domain_modified_count < 100,
  Freq(response_type),
  new_user_domain_modified_count
][order(new_user_domain_modified_count)]
p_session_types <- ggplot(
  logs_session[level != "correct"],
  aes(
    x = new_user_domain_modified_count,
    y = perc,
    group = level,
    color = level
  )
) +
  geom_line(linewidth = 1) +
  geom_point(aes(size = freq), alpha = 0.6) +
  scale_discrete_manual(
    aesthetics = c("fill", "color"),
    values = colors[c(2, 4)],
    name = "Response",
    labels = c("Error", "Question Mark")
  ) +
  labs(x = "Session Count", y = "Probability of Response", size = "Frequency") +
  scale_x_continuous(breaks = c(1, 25, 50, 75, 100), limits = c(1, 100)) +
  theme_bw() +
  theme(
    text = element_text(size = 14)
  )
p_session_types
save_plot(p_session_types, name = "response_types_over_sessions")
```

# Model analysis

To test whether problem-skipping is a latent trait distinguishable from accuracy, we compare four nested mixed-effects logistic regression models fitted on the long-format IRTree data (`dat_irt`). Each model represents a different hypothesis about the structure of individual differences. Model 1 (fully estimated) allows both student and item random effects to vary freely across nodes — this is the full IRTree. Models 2 and 3 constrain item and student effects respectively to be equal across nodes, and Model 4 (the unidimensional baseline) collapses both. If problem-skipping and accuracy are truly distinct, Model 1 should fit best on both in-sample (AIC/BIC) and out-of-sample (RMSE via k-fold cross-validation) criteria.

## Load tree model
```{r load-models}
load(file.path(results_dir, "models/mod_tree.Rdata"))
load(file.path(results_dir, "models/sum_mod_tree.Rdata"))
```

## Fixed effects

The fixed intercept in Model 1 captures the average log-odds of an incorrect response across all students and items, after accounting for the skip decision at Node 1. Converting this to a probability gives the baseline error rate in the data — the expected proportion of incorrect responses for a student and item both at the population mean.

```{r fixed-coef}
int <- mod_summary$coefficients[1]
1 / (1 + exp(-int))
```


## Inspect random effects

The random effects from Model 1 represent individual deviation from the population mean at each node. Node 1 effects capture how much each student or item deviates from the average skip propensity; Node 2 effects capture deviation from the average error propensity. These diagnostic plots check that the distributions are approximately normal and well-behaved, as assumed by the model.

```{r diagnostic-plots, eval = FALSE}
ranefs_mod <- ranef(mod_tree) 
# save(ranefs_mod, file = "ranefs_mod.Rdata")
# load("ranefs_mod.Rdata")
hist(
  plogis(ranefs_mod$user_id[, 1]),
  breaks = 50,
  main = "Distribution of random user effects, node 1"
)
hist(
  plogis(ranefs_mod$user_id[, 2]),
  breaks = 50,
  main = "Distribution of random user effects, node 2"
)
plot(
  plogis(ranefs_mod$user_id[, 1]),
  plogis(ranefs_mod$user_id[, 2]),
  xlab = "Random user effects, node 1",
  ylab = "Random user effects, node 2"
)

hist(
  plogis(ranefs_mod$item_id[, 1]),
  breaks = 50,
  main = "Distribution of random item effects, node 1"
)
hist(
  plogis(ranefs_mod$item_id[, 2]),
  breaks = 50,
  main = "Distribution of random item effects, node 2"
)
plot(
  plogis(ranefs_mod$item_id[, 1]),
  plogis(ranefs_mod$item_id[, 2]),
  xlab = "Random item effects, node 1",
  ylab = "Random item effects, node 2"
)
```

## Extract random effects

```{r ranefs}
ranef_user_node1 <- ranefs_mod$user_id[, 1]
ranef_user_node2 <- ranefs_mod$user_id[, 2]

ranef_item_node1 <- ranefs_mod$item_id[, 1]
ranef_item_node2 <- ranefs_mod$item_id[, 2]

ranefs <- data.frame(
  node = factor(c(
    rep(c(1, 2), each = length(ranef_user_node1)),
    rep(c(1, 2), each = length(ranef_item_node1))
  )),
  estimate = c(
    rep("user", times = length(ranef_user_node1) * 2),
    rep("item", times = length(ranef_item_node1) * 2)
  ),
  id = c(
    rep(rownames(ranefs_mod$user_id), 2),
    rep(rownames(ranefs_mod$item_id), 2)
  ),
  # values are converted to probabilities
  value = c(
    ranef_user_node1,
    ranef_user_node2,
    ranef_item_node1,
    ranef_item_node2
  )
)
rm(ranef_user_node1, ranef_user_node2, ranef_item_node1, ranef_item_node2)
# save(ranefs, file = "ranefs_df.Rdata")
# load("ranefs_df.Rdata")
```

## Distribution of random effects

The density plots below show the marginal distributions of the Node 1 and Node 2 random effects for both students and items. A well-fitting model should produce approximately symmetric, unimodal distributions centred near zero on the log-odds scale.

```{r plot-ranefs}
ggplot(ranefs, aes(x = value, fill = node, color = node)) +
  geom_density(alpha = 0.7) +
  scale_discrete_manual(
    aesthetics = c("fill", "color"),
    values = c("tan1", "tomato3")
  ) +
  facet_grid(node ~ estimate) +
  theme_bw()
```

## Correlation between random effects

The central results of RQ2. The estimated correlation between Node 1 and Node 2 random effects — separately for students (θ) and items (β) — indicates how strongly skip propensity and accuracy covary. A correlation of 1 would mean the two are perfectly redundant (the assumption of the unidimensional model); a correlation near 0 would mean they are entirely independent. The scatter plots display individual estimates; the regression line summarises the linear relationship; and the marginal density plots show the distribution of each parameter.

```{r plot-nodecor}
cor_plot_user <- ranefs %>%
  filter(estimate == "user") %>%
  pivot_wider(names_from = node, values_from = value) %>%
  ggplot(aes(x = `1`, y = `2`)) +
  geom_point(color = "gray", alpha = 0.3) +
  geom_smooth(method = "lm", color = "tan2", fill = "tan2") +
  labs(x = "Node 1", y = "Node 2") +
  theme_bw() +
  theme(legend.position = "none", text = element_text(size = 12))

cor_plot_item <- ranefs %>%
  filter(estimate == "item") %>%
  pivot_wider(names_from = node, values_from = value) %>%
  ggplot(aes(x = `1`, y = `2`)) +
  geom_point(color = "gray", alpha = 0.3) +
  geom_smooth(method = "lm", color = "tomato3", fill = "tomato3") +
  labs(x = "Node 1", y = "Node 2") +
  theme_bw() +
  theme(legend.position = "none", text = element_text(size = 12))

# add marginal dist
cor_plot_item_marginal <- ggMarginal(
  cor_plot_item + labs(x = "Item Skipping Threshold", y = "Item Difficulty"),
  type = "density",
  fill = "tomato3",
  alpha = 0.4,
  color = "tomato3"
)
cor_plot_user_marginal <- ggMarginal(
  cor_plot_user +
    labs(
      x = "User Propensity to Skip an Item",
      y = "User Propensity for Incorrect Response"
    ),
  type = "density",
  fill = "tan2",
  alpha = 0.4,
  color = "tan2"
)

combined_plot <- plot_grid(
  cor_plot_item_marginal,
  cor_plot_user_marginal,
  ncol = 2
)
rm(cor_plot_item, cor_plot_item_marginal, cor_plot_user_marginal)
combined_plot
save_plot(combined_plot, name = "plot_node_correlations", height = 4, width = 6)
```


## Individual response patterns

The correlation plot in the previous section summarises the average relationship between skip propensity and accuracy, but the scatter around this relationship is itself informative. Students who deviate from the average pattern — for example, a student with high accuracy but also high skipping — may be employing a qualitatively different strategy than students who conform to the trend. To explore this, we sample two students from each quadrant of the θ scatter plot (high/low on each node), visualise their full response sequences, and look for patterns in response type and timing that might reveal underlying strategies.

```{r ind-responses}
df_user <- data.frame()
ind_plots <- list()
quadrant_names <- c("UR", "LR", "LL", "UL")
quadrant_desc <- c(
  "UR" = ": Low Accuracy, High Skipping",
  "LR" = ": High Accuracy, High Skipping",
  "LL" = ": High Accuracy, Low Skipping",
  "UL" = ": Low Accuracy, Low Skipping"
)

for (i in seq_along(quadrant_names)) {
  # set boundary for which areas of correlation plot to sample from
  # upper right, lower right, lower left, upper left
  min_node1 <- c(1, 1, -4, -4)[i]
  max_node1 <- c(6, 6, -1, 0)[i]
  min_node2 <- c(1.5, -2, -2, 1)[i]
  max_node2 <- c(4, 0, 0, 4)[i]

  # sample 2 users per quadrant
  set.seed(random_seed)
  users <- sample(
    as.numeric(rownames(ranefs_mod$user_id[
      which(
        ranefs_mod$user_id$`factor(node)1` > min_node1 &
          ranefs_mod$user_id$`factor(node)1` < max_node1 &
          ranefs_mod$user_id$`factor(node)2` > min_node2 &
          ranefs_mod$user_id$`factor(node)2` < max_node2
      ),
    ])),
    2
  )

  # store sampled users; labels are assigned later based on clockwise position
  df_user <- rbind(
    df_user,
    data.frame(
      id = users,
      quadrant = quadrant_names[i]
    )
  )
}

users_df <- as.data.frame(ranefs_mod$user_id)
users_df$user_id <- as.numeric(rownames(users_df))
names(users_df)[1:2] <- c("node1", "node2")

df_user <- df_user %>%
  left_join(users_df, by = c("id" = "user_id")) %>%
  group_by(quadrant) %>%
  mutate(
    # Rank within each quadrant so ranks are always 1 or 2.
    # UR/LR: descending node2 so the topmost point in the scatter gets the
    #        lower rank (encountered first going clockwise down the right side).
    # LL/UL: ascending node2 so the bottommost point gets the lower rank
    #        (encountered first going clockwise up the left side).
    within_rank = case_when(
      quadrant %in% c("UR", "LR") ~ dplyr::min_rank(dplyr::desc(node2)),
      quadrant %in% c("LL", "UL") ~ dplyr::min_rank(node2)
    )
  ) %>%
  ungroup() %>%
  mutate(
    slot_order = case_when(
      quadrant == "UR" ~ within_rank,
      quadrant == "LR" ~ 2L + within_rank,
      quadrant == "LL" ~ 4L + within_rank,
      quadrant == "UL" ~ 6L + within_rank
    )
  ) %>%
  arrange(slot_order) %>%
  mutate(label = LETTERS[seq_len(n())])

for (quadrant in quadrant_names) {
  quadrant_users <- df_user %>%
    filter(quadrant == !!quadrant) %>%
    arrange(slot_order)

  users <- quadrant_users$id
  user_labels <- quadrant_users$label

  # prepare data for individual user response time plots
  dat_sub <- logs[logs$user_id %in% users, ]
  dat_sub$response_new <- ifelse(
    dat_sub$q == 1,
    "Question Mark",
    ifelse(dat_sub$correct_answered == 0, "Error", "Correct")
  )
  dat_sub <- dat_sub[order(dat_sub$id), ]
  dat_sub$rt <- dat_sub$rt / 1000   # convert ms → s
  dat_sub$item_number <- 1:nrow(dat_sub)
  # For left-column quadrants (LL, UL) reverse the factor levels so that the
  # lower-lettered user renders at the bottom of the panel, giving a
  # bottom-to-top reading of e→f (LL) and g→h (UL) that matches the
  # clockwise direction of the scatter plot labels.
  facet_levels <- if (quadrant %in% c("LL", "UL")) rev(users) else users
  dat_sub$user_id <- factor(dat_sub$user_id, levels = facet_levels)

  # facet labels
  custom_labels <- setNames(
    paste(
      user_labels,
      quadrant_desc[[quadrant]]
    ),
    users
  )

  # individual user response time plots
  p <- ggplot(
    dat_sub,
    aes(x = factor(item_number), y = rt, color = response_new)
  ) +
    geom_path(group = 1, color = "black", alpha = 0.3) +
    geom_point() +
    labs(x = "Item count", y = "Response Time", color = "Response") +
    scale_color_manual(
      values = c(
        "Correct" = "gray40",
        "Error" = "tomato3",
        "Question Mark" = "tan1"
      )
    ) +
    facet_wrap(
      ~user_id,
      nrow = 10,
      scales = "free",
      labeller = labeller(user_id = custom_labels)
    ) +
    scale_x_discrete(breaks = function(x) {
      x[seq(1, length(x), by = length(x) / 10)]
    }) +
    theme_bw()

  ind_plots[[quadrant]] <- p
}

p
```


```{r plot-ind-responses }
ranefs[ranefs$id %in% df_user$id, ]
df_ind <- merge(ranefs, df_user[, c("id", "label")], by = "id", all.x = TRUE)
df_ind <- df_ind %>%
  filter(estimate == "user") %>%
  pivot_wider(names_from = node, values_from = value)

df_ind_sub <- df_ind[complete.cases(df_ind$label), ]

cor_plot <- ggplot(df_ind, aes(x = `1`, y = `2`)) +
  geom_point(color = "gray40", alpha = 0.3) +
  geom_smooth(method = "lm", color = "gray7", fill = "gray7") +
  geom_point(data = df_ind_sub, aes(x = `1`, y = `2`), color = "tomato2") +
  geom_label(data = df_ind_sub, aes(label = label), size = 3, nudge_x = 0.2, nudge_y = 0.2) +
  labs(
    x = "User Propensity to Skip an Item",
    y = "User Propensity for Incorrect Response"
  ) +
  theme_bw() +
  theme(legend.position = "none", text = element_text(size = 12))

plotsAB <- plot_grid(
  ind_plots$UR +
    ylim(0, 20) +
    theme(legend.position = "none", axis.title = element_blank()),
  ind_plots$LR +
    ylim(0, 20) +
    theme(legend.position = "none", axis.title = element_blank()),
  ncol = 1
)

plotsCD <- plot_grid(
  ind_plots$UL +
    ylim(0, 20) +
    theme(legend.position = "none", axis.title = element_blank()),
  ind_plots$LL +
    ylim(0, 20) +
    theme(legend.position = "none", axis.title = element_blank()),
  ncol = 1
)

legend_plot <- ggplot(
  data.frame(
    x = c(1, 2, 3),
    y = c(1, 2, 3),
    label = c("Correct", "Incorrect", "Question Mark")
  ),
  aes(x = x, y = y, color = label)
) +
  geom_point(alpha = 0, size = 4) +
  scale_color_manual(
    name = "Response Type",
    values = c(
      "Correct" = "gray30",
      "Incorrect" = "tomato3",
      "Question Mark" = "tan2"
    ),
    labels = c("Correct", "Incorrect", "Question Mark")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5, alpha = 1))) + # Control legend appearance
  theme_void() + # Remove axes and background
  theme(legend.position = "bottom", text = element_text(size = 14))


full_grid = plot_grid(
  plotsCD,
  cor_plot + theme(plot.margin = unit(c(2, 0.5, 2, 0.5), "cm")),
  plotsAB,
  ncol = 4, # Adjust number of columns
  rel_widths = c(2, 1, 2, 0.5), # Set the width of the legend column
  rel_heights = c(2, 1, 2)
)

# Shared y-axis label for the four response-sequence panels.
# ggdraw() + draw_label() creates a plain rotated text grob; slotting it in as
# a narrow left column via plot_grid() keeps everything aligned without
# touching the individual sub-plots.
y_label <- ggdraw() +
  draw_label("Response Time (s)", angle = 90, size = 12)

full_grid_with_ylab <- plot_grid(
  y_label,
  full_grid,
  ncol = 2,
  rel_widths = c(0.04, 1)
)

full_grid_legend = plot_grid(
  full_grid_with_ylab,
  legend_plot + theme(plot.margin = unit(c(0, 2, 0, -1), "cm")),
  ncol = 1,
  rel_heights = c(12, 1)
)

full_grid_legend
#save_plot(full_grid_legend, name = "full_grid_legend", height = 6, width = 18)
```

```{r ind-responses-variations}
# Helper function: sample 8 users (2 per quadrant) using `seed`, build the
# full composite plot, and return it.  All objects it relies on
# (ranefs_mod, ranefs, logs, users_df, quadrant_names, quadrant_desc,
# legend_plot) are created in earlier chunks and available in the environment.
build_ind_grid <- function(seed) {

  df_user_v <- data.frame()

  set.seed(seed)
  for (i in seq_along(quadrant_names)) {
    min_node1 <- c(1, 1, -4, -4)[i]
    max_node1 <- c(6, 6, -1, 0)[i]
    min_node2 <- c(1.5, -2, -2, 1)[i]
    max_node2 <- c(4, 0, 0, 4)[i]

    users_v <- sample(
      as.numeric(rownames(ranefs_mod$user_id[
        which(
          ranefs_mod$user_id$`factor(node)1` > min_node1 &
            ranefs_mod$user_id$`factor(node)1` < max_node1 &
            ranefs_mod$user_id$`factor(node)2` > min_node2 &
            ranefs_mod$user_id$`factor(node)2` < max_node2
        ),
      ])),
      2
    )
    df_user_v <- rbind(
      df_user_v,
      data.frame(id = users_v, quadrant = quadrant_names[i])
    )
  }

  df_user_v <- df_user_v %>%
    left_join(users_df, by = c("id" = "user_id")) %>%
    group_by(quadrant) %>%
    mutate(
      within_rank = case_when(
        quadrant %in% c("UR", "LR") ~ dplyr::min_rank(dplyr::desc(node2)),
        quadrant %in% c("LL", "UL") ~ dplyr::min_rank(node2)
      )
    ) %>%
    ungroup() %>%
    mutate(
      slot_order = case_when(
        quadrant == "UR" ~ within_rank,
        quadrant == "LR" ~ 2L + within_rank,
        quadrant == "LL" ~ 4L + within_rank,
        quadrant == "UL" ~ 6L + within_rank
      )
    ) %>%
    arrange(slot_order) %>%
    mutate(label = LETTERS[seq_len(n())])

  ind_plots_v <- list()
  for (quadrant in quadrant_names) {
    quadrant_users_v <- df_user_v %>%
      filter(quadrant == !!quadrant) %>%
      arrange(slot_order)

    users_v      <- quadrant_users_v$id
    user_labels_v <- quadrant_users_v$label

    dat_sub_v <- logs[logs$user_id %in% users_v, ]
    dat_sub_v$response_new <- ifelse(
      dat_sub_v$q == 1,
      "Question Mark",
      ifelse(dat_sub_v$correct_answered == 0, "Error", "Correct")
    )
    dat_sub_v <- dat_sub_v[order(dat_sub_v$id), ]
    dat_sub_v$rt          <- dat_sub_v$rt / 1000  # ms -> s
    dat_sub_v$item_number <- 1:nrow(dat_sub_v)

    facet_levels_v <- if (quadrant %in% c("LL", "UL")) rev(users_v) else users_v
    dat_sub_v$user_id <- factor(dat_sub_v$user_id, levels = facet_levels_v)

    custom_labels_v <- setNames(
      paste(user_labels_v, quadrant_desc[[quadrant]]),
      users_v
    )

    ind_plots_v[[quadrant]] <- ggplot(
      dat_sub_v,
      aes(x = factor(item_number), y = rt, color = response_new)
    ) +
      geom_path(group = 1, color = "black", alpha = 0.3) +
      geom_point() +
      labs(x = "Item count", y = "Response Time (s)", color = "Response") +
      scale_color_manual(values = c(
        "Correct"       = "gray40",
        "Error"         = "tomato3",
        "Question Mark" = "tan1"
      )) +
      facet_wrap(
        ~user_id, nrow = 10, scales = "free",
        labeller = labeller(user_id = custom_labels_v)
      ) +
      scale_x_discrete(breaks = function(x) {
        x[seq(1, length(x), by = length(x) / 10)]
      }) +
      theme_bw()
  }

  # Scatter plot with the sampled users highlighted
  df_ind_v     <- merge(ranefs, df_user_v[, c("id", "label")], by = "id", all.x = TRUE)
  df_ind_v     <- df_ind_v %>%
    filter(estimate == "user") %>%
    pivot_wider(names_from = node, values_from = value)
  df_ind_sub_v <- df_ind_v[complete.cases(df_ind_v$label), ]

  cor_plot_v <- ggplot(df_ind_v, aes(x = `1`, y = `2`)) +
    geom_point(color = "gray40", alpha = 0.3) +
    geom_smooth(method = "lm", color = "gray7", fill = "gray7") +
    geom_point(data = df_ind_sub_v, aes(x = `1`, y = `2`), color = "tomato2") +
    geom_label(
      data = df_ind_sub_v, aes(label = label),
      size = 3, nudge_x = 0.2, nudge_y = 0.2
    ) +
    labs(
      x = "User Propensity to Skip an Item",
      y = "User Propensity for Incorrect Response"
    ) +
    theme_bw() +
    theme(legend.position = "none", text = element_text(size = 12))

  plotsAB_v <- plot_grid(
    ind_plots_v$UR + ylim(0, 20) +
      theme(legend.position = "none", axis.title = element_blank()),
    ind_plots_v$LR + ylim(0, 20) +
      theme(legend.position = "none", axis.title = element_blank()),
    ncol = 1
  )
  plotsCD_v <- plot_grid(
    ind_plots_v$UL + ylim(0, 20) +
      theme(legend.position = "none", axis.title = element_blank()),
    ind_plots_v$LL + ylim(0, 20) +
      theme(legend.position = "none", axis.title = element_blank()),
    ncol = 1
  )

  full_grid_v <- plot_grid(
    plotsCD_v,
    cor_plot_v + theme(plot.margin = unit(c(2, 0.5, 2, 0.5), "cm")),
    plotsAB_v,
    ncol = 4,
    rel_widths  = c(2, 1, 2, 0.5),
    rel_heights = c(2, 1, 2)
  )

  full_grid_with_ylab_v <- plot_grid(
    ggdraw() + draw_label("Response Time (s)", angle = 90, size = 12),
    full_grid_v,
    ncol = 2,
    rel_widths = c(0.04, 1)
  )

  plot_grid(
    full_grid_with_ylab_v,
    legend_plot + theme(plot.margin = unit(c(0, 2, 0, -1), "cm")),
    ncol = 1,
    rel_heights = c(12, 1)
  )
}

# Output subfolder (created automatically if absent)
ind_resp_dir <- file.path(plots_dir, "individual_responses")
dir.create(ind_resp_dir, showWarnings = FALSE, recursive = TRUE)

# Ten fixed seeds — change or extend as needed
variation_seeds <- c(42, 123, 456, 789, 1011, 2022, 3141, 5050, 7777, 9999)

for (k in seq_along(variation_seeds)) {
  grid_v <- build_ind_grid(variation_seeds[k])
  ggsave(
    filename = file.path(ind_resp_dir, sprintf("ind_responses_v%02d.png", k)),
    plot     = grid_v,
    height   = 6,
    width    = 18,
    bg       = "white"
  )
  message(sprintf("Saved version %d / %d (seed = %d)", k, length(variation_seeds), variation_seeds[k]))
}
```

The combined figure places each student's response sequence alongside their position in the θ scatter plot, allowing the reader to connect estimated latent parameters to observable behaviour. Students in the upper-right quadrant (high skipping, low accuracy) tend to show frequent, early skips interspersed with slow incorrect responses, consistent with low engagement or difficulty. Students in the lower-right quadrant (high skipping, high accuracy) are particularly informative: their sequences often reveal long runs of rapid skips followed by short bursts of correct responses, suggesting a deliberate strategy of avoiding effort on items perceived as unrewarding rather than unknown. This pattern of high skipping despite high accuracy is inconsistent with the assumptions of a unidimensional model, where skipping and inaccuracy are treated as manifestations of the same underlying trait.
